Lossy compression of DNA sequencing quality data

### 2015-04-10 manuscript/q-compression.tex
Various small changes in text and a figure legend. More to do.

\section{Motivation:}
The \textsc{fastq} file format has become the \emph{de facto} standard
for storing next-generation sequencing data, containing nucleotide
information along with a quantitative measure of the reliability of
individual base calls. As the cost of sequencing continues to
decrease, the rate of sequencing data production is increasing,
requiring efficient ways of storing and transferring this vast amount
of data. Most methods on sequencing data compression focus on
compressing nucleotide information without any loss of information.
Quality data, however, have different properties than nucleotide data,
and methods compressing nucleotide sequences efficiently do not
perform as well on quality sequences. Furthermore, while lossless
representation is necessary for nucleotide sequences, it is not an
essential requirement for quality values.

@@

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the
  \textit{Rhodobacter sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries, and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus musculus} data sets. 2B: 2-bin
  encoding. P64, P128, and P256: Quality profile compression with 64, 128, and 256
  profiles. R0-R7: Polynomial regression-based compression using 0-7
  degrees. Dashed line corresponds to lossless Bzip2 compression of
  quality values. \em}\label{fig:mse_vs_bpbp}
\end{figure*}

@@

As the order of the polynomial increases, the
bits/base-call increase and the \textsc{mse} decreases at an
exponential rate.
The 7th-order polynomial regression has the highest bits/base-call and in the \textit{Mus musculus} data set, and requires more storage than the \textsc{ascii} original quality values.  A 7th-order polynomial requires storing eight floating point values, resulting in 32 bytes per quality sequence.  The read length of the \textit{Mus musculus} data set is only 26, so the 7th-order polynomial regression is storing six more bytes than necessary to losslessly encode the quality data.

%\begin{figure}[!tpb]%figure3
%\centerline{\includegraphics[width=3.65in]{mse_short.png}}
%\caption{Mean squared error versus bits/base-call for for different
%  compression methods applied to the short-jump library from
%  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
%  encoding. P32 and P256: Quality profile compression with 32 and 256
%  profiles. R0-R5: Polynomial regression-based compression using 0-5
%  degrees. Dashed line corresponds to lossless bzip compression of
%  quality values. \em}\label{fig:mse_vs_bpbp_jump}
%\end{figure}

>>>>

\section{Motivation:}

The \textsc{fastq} file format has become the \emph{de facto} standard
for storing next-generation sequencing data, containing nucleotide
information along with a quantitative measure of the reliability of
individual base calls. As the cost of sequencing continues to
decrease, the rate of sequencing data production is increasing,
requiring efficient ways of storing and transferring this vast amount
of data. Most methods on sequencing data compression focus on
compressing nucleotide information without any loss of information.
Quality data, however, have different properties than nucleotide data,
and methods compressing nucleotide sequences efficiently do not
perform as well on quality sequences. Furthermore, while lossless
representation is necessary for nucleotide sequences, it is not an
essential requirement for quality values.

@@

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the \textit{Rhodobacter sphaeroides}
  2.4.1, and \textit{Homo sapiens} chromosome 14 fragment libraries,
  and \textit{Escherichia coli} str. K-12 MG1655, and \textit{Mus
    musculus} data sets. 2B --- 2-bin encoding; P$n$ 
  --- profiling with $n$ profiles; R$n$ --- modeling with polynomial regression
  models of degree $n$; Q$n$ --- QualComp with rate parameter 
  of $n$. Dashed line corresponds to lossless Bzip2
  compression of quality values, and dotted line corresponds to
  original uncompressed data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

As the order of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate.  The 7th-order
polynomial regression has the highest bits/base-call and in the
\textit{Mus musculus} data set, and requires more storage than the
\textsc{ascii} original quality values.  A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per quality sequence.  The read length of the \textit{Mus musculus}
data set is only 26, so the 7th-degree polynomial regression is storing
six more bytes than necessary for lossless encoding the quality data.
